Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2014-12-23 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Figure fixing.

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
\caption{M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.}
\label{fig:ADS_M-bias}
\end{figure}

@@

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\end{figure}

@@

Generally, this strategy is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads. However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101. Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded.

@@

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

\begin{figure}[h!]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
\label{fig:methtuple_example1}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
\label{fig:methtuple_runtime}
\end{figure}

>>>>

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/ADS_M-bias.pdf}
\caption{M-bias plot for the ADS sample from the Lister (2011) dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. We see significant CpG M-bias at the start of read 2 and some noise at the start of read 1.}
\label{fig:ADS_M-bias}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/E18VA_M-bias.pdf}
\caption{M-bias plot for the E18VA sample from the EPISCOPE dataset. Each of read 1 (R1) and read 2 (R2) are plotted separately. For CpGs in read 1, we see noise at the start of the read, followed by a downward slope in the M-bias, which ends with a spike. For CpGs in read 2, we see a downward spike at the start of the read following by a gradual increase in the M-bias curve, with a spike at read-position 101 and a spike at the last read positions for all methylation types. The spike at read-position 101 is also evidence, albeit to a lesser extent, in read 1. This position should be ignored in downstream analyses but we do not necessarily also want to ignore read-positions $102-150$ since this would remove one-third of the data.}
\end{figure}

@@

Generally, this strategy is okay because M-bias tends to occur as runs of read-positions at the 5' and 3' ends of reads. However, occasionally there are spikes in the M-bias plot, which indicate specific read-positions that we would like to exclude. Figure \ref{fig:E18VA_M-bias} shows an example of such a spike that occurred at read-position 101. Using `bismark_methylation_extractor` we would be forced to either retain this position or to ignore read-positions $101-150$, effectively ignoring one third of the sequencing data, much of it unaffected by M-bias. My `methtuple` software (Section \ref{sec:methtuple}) avoids the unnecessary exclusion of those bases by allowing the user to specify the exact read-positions that she wants excluded, however, this is not without its own problems, as we shall see in Section \ref{sec:cometh_nil_geq0}.

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=\textwidth]{../figures/Bis-SNP.pdf}
\caption{texttt{Bis-SNP} is able to distinguish unmethylated cytosines (site 1), from cytosine to thymine genetic variants (site 2) and thymine to (unmethylated) cytosine genetic variants (site 3) by examining the reads mapped to the reverse strand. For all three loci, the reads mapped to the forward strand contain a thymine. However, it is the base on the reverse strand that reveal the true genotype. When combined with the reference genome it can be inferred whether the sample's genome, which isn't directly observed, has a genetic variant at that location. This is only possible with bisulfite-data generated using the directional protocol. This figure is adapted from \citet{Liu:2012ge}.}
\label{fig:Bis-SNP}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\centering
\includegraphics[width=0.6\textwidth]{../figures/methtuple_example1.pdf}
\caption{Diagram of three-paired end reads --- \texttt{A}, \texttt{B} and \texttt{C} --- mapping to a region containing five methylation loci --- \texttt{1} $-$ \texttt{5}.}
\label{fig:methtuple_example1}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example2.pdf}
\caption{1-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example2}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example3.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2}.}
\label{fig:methtuple_example3}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_example4.pdf}
\caption{3-tuples produced for each read for the toy example in Figure \ref{fig:methtuple_example2} when the \texttt{--all-combinations} flag is set.}
\label{fig:methtuple_example4}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Lister_CpGs_per_read.pdf}
\caption{Number of CpGs per read for the Lister dataset.}
\label{fig:Lister_CpGs_per_read}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/3-tuples.pdf}
\caption{Example output of \texttt{methtuple} for 3-tuples.}
\label{fig:3-tuples}
\end{figure}

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{--all-combinations} was set. Note that the reported $49$ samples separately counts each of the Ziller sequencing runs.}
\label{fig:methtuple_runtime}
\end{figure}
